---
title: "Analysis: Baltimore city salaries, FY 2019"
author: "Christine"
date: "10/03/2019"
output:   
  md_document:
    variant: markdown_github
---

## Analysis: Baltimore city salaries, fiscal year 2019

The Baltimore Sun analyzed data from [Open Baltimore](https://data.baltimorecity.gov/City-Government/Baltimore-City-Employee-Salaries-FY2019/6xv6-e66h) on salaries for city employees employed on June 30, 2019. The dataset captures salary from July 1, 2018 through June 30, 2019.

Here are the key statistics reported in the story:

- Sgt. Ethan Newberg, a 24-year veteran of the city’s police force, earned $260,775, more than double his base salary of $107,807
- In the previous fiscal year, Newberg ranked second on the list of best-paid city employees, making $243,132
- Police Commissioner Michael Harrison and his two deputy commissioners, Danny Murphy and Michael Sullivan, were the only police employees whose base salary ranked in the top 10
- Seven of the 10 highest-paid city employees were police officers
- Harrison makes a base salary of $275,000. Former police commissioner Darryl DeSousa made an annual base salary of $210,000
- Total police department pay exceeded base salary by more than $40 million, down from 2018, when the difference was $51 million 
- The median pay for an employee in the police department was $93,278
- The fire department had the second-highest total pay, with total compensation exceeding base salaries by more than $16 million
- The median pay for a fire department employee was $82,097
- The second-highest paid employee was Frank Johnson, who earned $251,922 on a salary of $250,000. The next-highest paid was William Harris Jr., a police sergeant who made $249,356 on a salary of $107,364. Harris was followed by police Lt. Thomas Mistysyn Jr., who earned $247,182 on a base salary of $122,049 and fire department employee Anthony Smith, who earned $243,747 on a base salary of $105,000. State’s Attorney Marilyn Mosby was the sixth-highest paid, with a salary of $238,772
- Harris was the city’s highest paid employee in fiscal year 2018
- Mayor Bernard C. “Jack” Young earned $127,984 on a salary of $184,832, nearly $10,000 more than what he earned as City Council President during the last fiscal year

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load R libraries

```{r, warning=F, message=F}
library('tidyverse')
library('janitor')
library('lubridate')
```

### Read salaries data
```{r, warning=F, message=F}
salaries.19 <- read_csv('input/Baltimore_City_Employee_Salaries_FY2019.csv') %>% clean_names()
salaries.18 <- read_csv('input/Baltimore_City_Employee_Salaries_FY2018.csv') %>% clean_names()
```

### Finding: Sgt. Ethan Newberg, a 24-year veteran of the city’s police force, earned $260,775, more than double his base salary of $107,807

Arrange the data by `gross` — the total compensation in fiscal year 2019 — and use `head()` to see the top salaries. Newberg is first.

```{r}
salaries.19 %>% arrange(desc(gross)) %>% head()
```

```{r}
print(paste(salaries.19[order(-salaries.19$gross), ][1,]$name, "was the highest earner in fiscal year 2019, making",
      salaries.19[order(-salaries.19$gross), ][1,]$gross, "or", 
      round(salaries.19[order(-salaries.19$gross), ][1,]$gross/salaries.19[order(-salaries.19$gross), ][1,]$annual_rt, 1), "times base pay."))
```

```{r}
print(paste("The highest earner has worked in the city for", round(as.numeric(difftime(today(), 
                                                                                  as.Date(mdy_hms(salaries.19[order(-salaries.19$gross), ][1,]$hire_dt)), unit="weeks")/52.25), 1), "years."))
```

### Finding: In the previous fiscal year, Newberg ranked second on the list of best-paid city employees, making $243,132

```{r}
salaries.18 %>% arrange(desc(gross)) %>% head()
```

### Finding: Police Commissioner Michael Harrison and his two deputy commissioners, Danny Murphy and Michael Sullivan, were the only police employees whose base salary ranked in the top 10

Arrange the data by `annual_rt` — the base salary in fiscal year 2019 — and use `head(10)` to see the top 10 salaries.

```{r}
salaries.19 %>% arrange(desc(annual_rt)) %>% head(10)
```

### Finding: Seven of the 10 highest-paid city employees were police officers

Create a column, `police`, that = 1 if the employee works in the police department; 0 if not. Arrange the top 10 by gross and sum the number of police.

```{r}
salaries.19 %>% 
  mutate(police = ifelse(grepl("Police", descr), 1, 0)) %>% 
  arrange(desc(gross)) %>% 
  head(10) %>% 
  summarise(police_in_top_10 = sum(police))
```

### Finding: Harrison’s base salary is $275,000 a year. Former police commissioner Darryl DeSousa made an annual base salary of $210,000

Use the 2019 and 2018 salary database to compare Harrison and DeSousa's pay

```{r}
print(salaries.19 %>% filter(jobtitle == 'Police Commissioner'))
print(salaries.18 %>% filter(jobtitle == 'Police Commissioner'))
```

### Finding: Total police department pay exceeded base salary by more than $40 million, down from 2018, when the difference was $52 million

Calculate the difference betwewen total base salary and actual pay for the police department in 2019 verus 2018

```{r}
salaries.19 %>% 
  filter(grepl("Police Department", descr)) %>% 
  summarise(base = sum(annual_rt, na.rm = T),
            actual = sum(gross, na.rm = T)) %>%
  mutate(difference = actual - base)
```
```{r}
salaries.18 %>% 
  filter(grepl("Police", descr)) %>% 
  summarise(base = sum(annual_rt, na.rm = T),
            actual = sum(gross, na.rm = T)) %>%
  mutate(difference = actual - base)
```

### Finding: The median pay for an employee in the police department was $93,278

Filter and summarize to calculate the median police department (including overtime) pay in 2019

```{r}
salaries.19 %>% 
  filter(grepl("Police", descr)) %>% 
  summarise(median = median(gross, na.rm = T)) 
```

### Finding: The fire department had the second-highest total pay, with salaries exceeding base salaries by more than $16 million

Group by department (remove the parenthetical numbers in the department `descr` column to get an accurate tally), and arrange by total gross pay. Use `head()` to see the top departments.
```{r, warning=F}
salaries.19.dept <- separate(salaries.19,
                           descr,
                           into = c('dept', 'number'),
                           sep = "\\(",
                           remove = F) %>%
  select(-number) %>% 
  mutate(dept = trimws(dept)) %>%
  group_by(dept) %>%
  summarise(actual = sum(gross, na.rm = T)) %>%
  arrange(desc(actual)) %>%
  head(10)

salaries.19.dept

```

The fire department is #2 after the police department.

Calculate total pay plus the difference between base and actual for the fire department.

```{r}
salaries.19 %>% 
  filter(grepl("Fire", descr)) %>% 
  summarise(base = sum(annual_rt, na.rm = T),
            actual = sum(gross, na.rm = T)) %>%
  mutate(difference = actual - base)
```
The difference is more than 16 million.

### Finding: The median pay for a fire department employee was $82,097

```{r}
salaries.19 %>% 
  filter(grepl("Fire", descr)) %>% 
  summarise(median = median(gross, na.rm = T)) 
```

### Finding: The second-highest paid employee was Frank Johnson, who earned $251,922 on a salary of $250,000. The next-highest paid was William Harris Jr., a police sergeant who made $249,356 on a salary of $107,364. Harris was followed by police Lt. Thomas Mistysyn Jr., who earned $247,182 on a base salary of $122,049 and fire department employee Anthony Smith, who earned $243,747 on a base salary of $105,000

Arrange the data by `gross` and use `head()` to see the top salaries

```{r}
salaries.19 %>% arrange(desc(gross)) %>% head()
```

### Finding: Harris was the city’s highest paid employee in fiscal year 2018

Arrange the 2018 data by `gross` and use `head(1)` to see the top paid employee

```{r}
salaries.18 %>% arrange(desc(gross)) %>% head(1)
```

### Finding: Mayor Bernard C. “Jack” Young earned $127,984 on a salary of $184,832, nearly $10,000 more than what he earned as City Council President during the last fiscal year

Filter the 2018 and 2019 data to find Young's salaries in those fiscal years

```{r}
print(salaries.18 %>% filter(name =='Young,Bernard C'))
print(salaries.19 %>% filter(name =='Young,Bernard C'))
```

```{r}
print(paste("Mayor Young earned $", 
      salaries.19[salaries.19$name =='Young,Bernard C', ]$gross - salaries.18[salaries.18$name =='Young,Bernard C',]$gross,
      "more in fiscal year 2019 than in fiscal year 2018."))
```

